### 2015-08-10 corral.tex
fixes

\corral is inherently unsound (i.e.\ can miss real races), because it performs bounded verification. However, \corral is precise and, assuming a precise environmental model, it will only report true races. \whoop takes advantage of this precision to report only feasible data races. Also note that our instrumentation conveniently tolerates some benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.

@@

\noindent\textbf{Sound Partial-Order Reduction }
%
By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} after each shared memory access of each entry point, and after every lock and unlock operation.\footnote{We acknowledge that in the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound. Our approach to accelerating \corral is simple and yet effective: if, thanks to \whoop's analysis, we know that a given statement that accesses shared memory cannot be involved in a data race, then we do not instrument a \texttt{yield} after this statement.  This is a form of \emph{partial order reduction}~\cite{DBLP:books/sp/Godefroid96}, and reduces the number of context-switches that \corral must consider in a \emph{sound} manner: there is no impact on the bugs that will be detected.  This is because each access for which a \texttt{yield} is suppressed is guaranteed to be protected by some lock (a consequence of lockset analysis).  If the access is a write, its effects are not visible by the other entry point in the pair until the lock is released.  If the access is a read, the value of the shared location cannot change until the lock is released.  The fact that a \texttt{yield} is placed after each unlock operation suffices to take account of communication between entry points via the shared memory location.

>>>>

\corral is inherently unsound (i.e.\ can miss real races), because it performs bounded verification. However, \corral is precise and, assuming a precise environmental model, it will only report true races. \whoop takes advantage of this precision to report only feasible data races.
%
Note that our instrumentation conveniently tolerates some benign races: it does not report a read-write race if the write access updates the racy memory location with the same value that already existed; it also does not report a write-write race if the two write accesses update the racy memory location with the same value, which can be different from the pre-existing.

@@

\noindent\textbf{Sound Partial-Order Reduction }
%
By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} after each shared memory access of each entry point, and after every lock and unlock operation.\footnote{We acknowledge that in the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound.
%
Our approach to accelerating \corral is simple and yet effective: if, thanks to \whoop's analysis, we know that a given statement that accesses shared memory cannot be involved in a data race, then we do not instrument a \texttt{yield} after this statement, and we also omit the \texttt{assert} that would check for a race.
%
This is a form of \emph{partial order reduction}~\cite{DBLP:books/sp/Godefroid96}, and reduces the number of context-switches that \corral must consider in a \emph{sound} manner: there is no impact on the bugs that will be detected.  This is because each access for which a \texttt{yield} is suppressed is guaranteed to be protected by some lock (a consequence of lockset analysis).  If the access is a write, its effects are not visible by the other entry point in the pair until the lock is released.  If the access is a read, the value of the shared location cannot change until the lock is released.  The fact that a \texttt{yield} is placed after each unlock operation suffices to take account of communication between entry points via the shared memory location.

### 2015-08-10 evaluation.tex
some fixes
fixes

For reproducibility, we make all of our tools (including source) and benchmarks available online:

\begin{scriptsize}
\textbf{\texttt{\url{http://multicore.doc.ic.ac.uk/tools/Whoop/ASE2015}}}
\end{scriptsize}

@@

Most of the races that \whoop and \corral discovered fall into two categories. The first is about accessing a global counter (or flag) from concurrent entry points, without holding a lock. This might be for performance, and indeed a lot of the races we found might be benign. Even benign races, though, lead to undefined behavior according to the C standard\COULDCUT{, and it is well known that undefined behaviors can lead to unexpected results when combined with aggressive compiler optimizations}. The second is about an entry point modifying a field of a struct (either global or passed as a parameter) without holding a lock. This can lead to a race if another entry point simultaneously accesses the same field of the same struct.

As an example of the second category, we found the following race in the generic\_nvram driver (see Fig.~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without holding a lock (\texttt{file} is passed as a parameter to \texttt{llseek}). This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.  We observed that another char driver, using the same APIs, \emph{does} use a mutex to protect the offset access in its \texttt{llseek} entry point, leading us to suspect that the race we found in generic\_nvram is not benign.  We have filed a bug report in the Linux kernel bug tracker and are awaiting a response.

>>>>

%For reproducibility, we make all of our tools (including source) and benchmarks available online:

%\begin{scriptsize}
%\textbf{\texttt{\url{http://multicore.doc.ic.ac.uk/tools/Whoop/ASE2015}}}
%\end{scriptsize}

@@

Most of the races that \whoop and \corral discovered fall into two categories. The first is about accessing a global counter (or flag) from concurrent entry points, without holding a lock. This might be for performance, and indeed a lot of the races we found might be benign. Even benign races, though, lead to undefined behavior according to the C standard.
%
%, and it is well known that undefined behaviors can lead to unexpected results when combined with aggressive compiler optimizations.
%
The second is about an entry point modifying a field of a struct (either global or passed as a parameter) without holding a lock. This can lead to a race if another entry point simultaneously accesses the same field of the same struct.

As an example of the second category, we found the following race in the generic\_nvram driver (see Fig.~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without holding a lock (\texttt{file} is passed as a parameter to \texttt{llseek}). This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.  We observed that another char driver, using the same APIs, \emph{does} use a mutex to protect the offset access in its \texttt{llseek} entry point, leading us to suspect that the race we found in generic\_nvram is not benign.

### 2015-08-10 intro.tex
fixes

Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software to provide an interface between the operating system and hardware devices that are attached to a computer. Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}. This has a negative impact on hardware product releases, as time-to-market is commonly dominated by driver development, verification, and validation schedules~\cite{yavatkar2012era}.
%
Even after a driver has shipped, it typically has many undetected errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel, while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers. Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} established that they account for 19\% of the total bugs in Linux drivers, showcasing their significance. The majority of these concurrency bugs were \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.

>>>>

Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software to provide an interface between the operating system and the devices that are attached to a computer. Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}. This has a negative impact on hardware product releases, as time-to-market is commonly dominated by driver development, verification, and validation schedules~\cite{yavatkar2012era}.
%
Even after a driver has shipped, it typically has many undetected errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel, while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers. Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} established that they account for 19\% of the total bugs in Linux drivers, showcasing their significance. The majority of these concurrency bugs were \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.

### 2015-08-10 whoop.tex
fixes

A thread terminates if it reaches the end of $\mainforthread{T}$, in which case its state is removed from the list of thread states.  Since our interest is in analysis of device drivers, which are reactive, we do not consider the notion of global program termination.

@@

The simple concurrent programming model of \S\ref{sec:symbolicpairwise} is deliberately idealistic to make it easy to describe our symbolic verification technique. In practice, Linux drivers are written in C, our technique does not know up-front which are the driver entry points, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers. We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop,\footnote{\url{https://github.com/pdeligia/whoop}} a practical, automated tool for detecting data races in drivers.

\noindent\textbf{Architecture }
%
Fig.~\ref{fig:whoop} depicts the \whoop toolchain. The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races. Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur,\footnote{\url{https://github.com/mc-imperial/chauffeur}} Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.  Boogie is a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ the Boogie verifier and \corral). Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Fig.~\ref{fig:whoop}--B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine. After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Fig.~\ref{fig:whoop}--C and \S\ref{corral}).

We engineered the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Fig.~\ref{fig:whoop}).  For the remaining components, we were able to reuse industrial-strength tools that are robust and battle-proven via their use in many complex software projects.

@@

\noindent\textbf{Translation for Verification }
%
Next, the driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form. Function calls (e.g.\ for locking and unlocking) are preserved in this translation and, hence, we do not need to keep track of them separately.
%
SMACK then translates the driver from LLVM-IR to Boogie IVL, which is the input language of \whoop. An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL. Thus, \whoop does not need to directly deal with pointers and alias analysis\COULDCUT{, a hard problem in its own right, which allows us to reuse robust existing techniques and focus instead on verification efforts}.

@@

\COULDCUT{
In practice, the candidates are generated at the Boogie IVL level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race-checking instrumentation.}

@@

Statically analyzing drivers requires ``closing'' the environment by abstracting away the low-level implementation details. To this end, we developed a simple model for the Linux kernel that consists of (nondeterministic) stub functions. A limitation of our model is that it can, and will, ultimately result in false positives. However, we currently only focus on finding data races, and thus can get away with over-approximating large parts of the Linux kernel, without losing too much precision. Making our model more precise is an ongoing manual effort that requires Linux kernel expertise. We argue that further work on the model is orthogonal to the contributions of this paper. Also, even if our symbolic analysis results in false positives, \whoop can still use the results to significantly speedup a more precise bug-finder, as seen in \S\ref{corral} and \S\ref{evaluation}.

>>>>

A thread terminates if it reaches the end of $\mainforthread{T}$, in which case its state is removed from the list of thread states.  Since our interest is in analysis of device drivers, which are reactive programs, we do not consider the notion of global program termination.

@@

The simple concurrent programming model of \S\ref{sec:symbolicpairwise} is deliberately idealistic to make it easy to describe our symbolic verification technique. In practice, Linux drivers are written in C, our technique does not know up-front which are the driver entry points, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers. We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop, a practical, automated tool for detecting data races in drivers.

\noindent\textbf{Architecture }
%
Fig.~\ref{fig:whoop} depicts the \whoop toolchain. The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races. Initially, \whoop uses three LLVM-based tools~\cite{lattner2004llvm}, Chauffeur, Clang and SMACK~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.  Boogie is a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ the Boogie verifier and \corral). Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Fig.~\ref{fig:whoop}--B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine. After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Fig.~\ref{fig:whoop}--C and \S\ref{corral}).

We engineered the Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}} and \whoop\footnote{\url{https://github.com/pdeligia/whoop}} components of our toolchain (denoted with grey boxes in Fig.~\ref{fig:whoop}).  For the remaining components, we were able to reuse industrial-strength tools that are robust and battle-proven via their use in many complex software projects.

@@

\noindent\textbf{Translation for Verification }
%
Next, the driver is compiled by Clang to LLVM-IR, a low-level assembly-like language in single static assignment (SSA) form. Function calls (e.g.\ for locking and unlocking) are preserved in this translation and, hence, we do not need to keep track of them separately.
%
SMACK then translates the driver from LLVM-IR to Boogie IVL, which is the input language of \whoop. An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL. Thus, \whoop does not need to directly deal with pointers and alias analysis.
%, a hard problem in its own right, which allows us to reuse robust existing techniques and focus instead on verification efforts.

@@

%In practice, the candidates are generated at the Boogie IVL level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race-checking instrumentation.

@@

Statically analyzing drivers requires ``closing'' the environment by abstracting away the low-level implementation details. To this end, we developed a simple model for the Linux kernel that consists of (nondeterministic) stub functions.
%
A limitation of our model is that it can be inaccurate, leading to semantic mismatches that can in turn lead to false positives and/or false negatives.
%
However, we currently only focus on finding data races, and thus can get away with over-approximating large parts of the Linux kernel, without losing too much precision. Making our model more precise is an ongoing manual effort that requires Linux kernel expertise. We argue that further work on the model is orthogonal to the contributions of this paper. Also, even if our symbolic analysis results in false positives, \whoop can still use the results to significantly speedup a more precise bug-finder, as seen in \S\ref{corral} and \S\ref{evaluation}.
